Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2015-12-19 threshold/draft.Rnw
Fixed numerous typos

Let $A=U\Sigma V^T$ be a singular value decomposition of $A$, where
$U\in\mathcal{R}^{m\times k}$, $V\in\mathcal{R}^{n\times k}$,
$U^TU = V^TV = I$,
and $\Sigma\in\mathcal{R}^{k\times k}$ is a diagonal matrix with
diagonal entries $\sigma_1 \ge \sigma_2 \ge \cdots \ge \sigma_k > 0$.  For
later convenience, we write $s_{1:p}=[\sigma_1, \sigma_2, \ldots, \sigma_p]^T$
to be the vector of the first $p\le k$ singular values along the diagonal of
$\Sigma$.  Denote the columns of $V$ as $V=[v_1, v_2, \ldots, v_k]$, where each
$v_j\in\mathcal{R}^n$, $j=1,2,\ldots,k$.  Then $\mathrm{cor}(A) = A^TA =
V\Sigma^2 V^T$ is a symmetric eigenvalue decomposition of the correlation
matrix with nonzero eigenvalues $\sigma_1^2, \sigma_2^2, \ldots, \sigma_k^2$.
The span of the columns of $V$ form an orthonormal basis of the range of the
correlation matrix $A^TA$.

@@

Pairs of vectors below the correlation threshold that lie more than one index
difference apart relative to the permutation $P$ can be similarly pruned by
replacing $D^1$ with $D^j$, $j=2,3,\ldots,\ell$ in the above expressions.
Following this process, we can produce a well-pruned candidate set of vector
pairs that contains all pairs that meet a given correlation threshold using
$\ell$ matrix vector products with matrices of order $n \times p$, where $p$ is
chosen in practice such that $p<<k$. Setting $p=10$ for the example shown in
Figure \ref{fig1} prunes all but $431$ possible vector pairs out of a total
$419(419-1)/2 = 87,571$ with $\ell=138$.

@@

We can relax the column mean and scale assumptions without introducing much
additional computational or storage overhead as follows. Assume $A$ is an
$m\times n$ real-valued matrix without any constant-valued columns. Let
$z=[z_1,z_2,\ldots,z_n]\in\mathcal{R}^n$ represent the vector of column means
of the $m\times n$ matrix $A$, $w=[w_1^2,w_2^2,\ldots,w_n^2]\in\mathcal{R}^n$
be the vector of squared column norms of $A$, and $W\in\mathcal{R}^{n\times n}$
a diagonal matrix with diagonal entries $1/\sqrt{w_i^2 - m z_i^2}$.  Let
$e\in\mathcal{R}^m$ be a vector of all ones.  Then $(A - ez^T)W$ is a centered
matrix with zero column means and scaled to have unit column norms, and
\[
\mathrm{cor}(A) = W^T (A-ez^T)^T (A-ez^T) W.
\]
The IRLBA, based on the Lanczos process, is a method of iterated matrix-vector
products. The main idea behind efficient application of IRLB to correlation
problems replaces matrix-vector products of the form $Ax$ with $AWx - ez^TWx$ in
the iterations, implicitly working with a scaled and centered matrix without
forming it.  This comes at the cost of storing two additional length $n$ vectors and
at most computing two additional vector inner products per matrix-vector product.
The \proglang{R} implementation \citep{irlbar} of IRLBA includes arguments for centering and
scaling the input matrix.

@@

We compare the {\tt tcor()} function with brute-force Pearson correlation
matrix computation computed in the fastest way we can, directly using matrix
multiplication instead of using the slower {\tt cor()} function.

@@

Anyone can quickly run the example to experiment with the algorithm and
compare its output with brute force results. The example uses the {\tt
EisenYeast} data from the {\tt biclust} \proglang{R} package \citep{biclust}.
The data consist of 80 experiments involving the
Saccharomyces Cerevisiae yeast measuring gene expression data across 6,221
Affymetrix probes, represented as a $80\times 6221$ real-valued matrix.  The
example finds all pairs of columns (gene expression measurements) that meet or exceed a
Pearson correlation value of 0.95.  We use a projected dimension $p=10$
in {\tt tcor()}.
\begin{small}
<<eval=TRUE,echo=TRUE>>=
library(tcor)
library(doMC)     # local parallel computation
library(biclust)  # for the BicatYeast data
data(EisenYeast)  # in gene by sample orientation
A = t(EisenYeast) # to correlate genes to genes
threshold = 0.95

@@

<<echo=FALSE, results=tex>>=
require(xtable)
ans = data.frame(Method=c("tcor","brute force"), `Wall clock time (s)`=c(t1, t2), `Peak excess memory use (MB)`=c(m1, m2), check.names=FALSE)
print(
    xtable(ans,
        caption = "EisenYeast results, threshold=0.95", label = "EisenYeast", caption.placement = "top")
  ,size="small", table.placement="ht", include.rownames=FALSE)
@

@@

Performance benefits from the {\tt tcor()} function become apparent with more
data.  The following example uses calculated beta values from the Illumina
Human Methylation450 BeadChip of 80 Adenoid Cystic Carcinoma (ACC) tumor
samples obtained from the Broad GDAC Cancer Genome Atlas dashboard \citep{gdac}.
See the \proglang{R} package implementation \citep{sup} for examples of
efficiently converting the downloaded sample methylation data into an R matrix.
We chose these data for their public availability as a general example of
computing thresholded correlation among sets of large numbers of vectors,
rather than for any specific biological meaning.

We removed 91,563 columns with all constant or missing values from the data,
leaving an $80\times 394014$ matrix. The example computes pairs of methylation
beta value column vectors with correlation values of 0.99 or more. We used a
{\tt tcor} projection dimension of 10 and the same performance measurement
methodology used in the first example. Despite its low rank and modest data
size, the problem is too large to compute using a na\"ive brute force
correlation method.

Our simple brute force implementation was not possible due to memory constraints.
We remark that the matrix multiplication central to the brute force
implementation can be decomposed into a sequence of smaller steps to
fit within given memory constraints. However, the brute force algorithm would still
need to evaluate all of the nearly 160 billion possible vector pair correlation
values.

@@

We present a simple algorithm and reference \proglang{R} implementation based on a
truncated singular value decomposition (SVD) that efficiently computes
thresholded subsets of large correlation and distance matrices.  We show that
the natural connection between data covariance and the SVD can result in
substantial computational and memory savings, and illustrate these savings with
a few genomics data experiments.  Moreover, the main computational work in the
algorithm is easily parallelized and easily deployed across computational
clusters.

>>>>

Let $A=U\Sigma V^T$ be a singular value decomposition of $A$, where
$U\in\mathcal{R}^{m\times k}$, $V\in\mathcal{R}^{n\times k}$,
$U^TU = V^TV = I$,
and $\Sigma\in\mathcal{R}^{k\times k}$ is a diagonal matrix with
diagonal entries $\sigma_1 \ge \sigma_2 \ge \cdots \ge \sigma_k > 0$.  For
later convenience, we write $s_{1:p}=[\sigma_1, \sigma_2, \ldots, \sigma_p]^T$
to be the vector of the first $p\le k$ singular values along the diagonal of
$\Sigma$.  Denote the columns of $V$ as $V=[v_1, v_2, \ldots, v_k]$, where each
$v_j\in\mathcal{R}^n$, $j=1,2,\ldots,k$.  Then $\mathrm{cor}(A) = A^TA =
V\Sigma^2 V^T$ is a symmetric eigenvalue decomposition of the correlation
matrix with nonzero eigenvalues $\sigma_1^2, \sigma_2^2, \ldots, \sigma_k^2$.
The columns of $V$ form an orthonormal basis of the range of the
correlation matrix $A^TA$.

@@

Pairs of vectors below the correlation threshold that lie more than one index
difference apart relative to the permutation $P$ can be similarly pruned by
replacing $D^1$ with $D^j$, $j=2,3,\ldots,\ell$ in the above expressions.
Following this process, we can produce a well-pruned candidate set of vector
pairs that contains all pairs that meet a given correlation threshold using
$\ell$ matrix vector products with matrices of order $n \times p$, where $p$ is
chosen in practice such that $p<<k$. Setting $p=10$ for the example shown in
Figure \ref{fig1} prunes all but $12,924$ possible vector pairs out of a total
$6221(6221 - 1) = 193,47,310$ with $\ell=1,715$.

@@

We can relax the column mean and scale assumptions without introducing much
additional computational or storage overhead as follows. Assume $A$ is an
$m\times n$ real-valued matrix without any constant-valued columns. Let
$z=[z_1,z_2,\ldots,z_n]\in\mathcal{R}^n$ represent the vector of column means
of $A$, $w=[w_1^2,w_2^2,\ldots,w_n^2]\in\mathcal{R}^n$
be the vector of squared column norms of $A$, and $W\in\mathcal{R}^{n\times n}$
a diagonal matrix with diagonal entries $1/\sqrt{w_i^2 - m z_i^2}$.  Let
$e\in\mathcal{R}^m$ be a vector of all ones.  Then $(A - ez^T)W$ is a centered
matrix with zero column means and scaled to have unit column norms, and
\[
\mathrm{cor}(A) = W^T (A-ez^T)^T (A-ez^T) W.
\]
The IRLBA, based on the Lanczos process, is a method of iterated matrix-vector
products. The main idea behind efficient application of IRLB to correlation
problems replaces matrix-vector products of the form $Ax$ with $AWx - ez^TWx$ in
the iterations, implicitly working with a scaled and centered matrix without
forming it.  This comes at the cost of storing two additional length $n$ vectors and
at most computing two additional vector inner products per matrix-vector product.
The \proglang{R} implementation \citep{irlbar} of IRLBA includes arguments for centering and
scaling the input matrix.

@@

We compare the {\tt tcor()} function with brute-force Pearson correlation
matrix computation computed in the fastest way we can, directly using matrix
multiplication instead of using the more general {\tt cor()} function.

@@

Anyone can quickly run the example to experiment with the algorithm and
compare its output with brute force results. The example uses the {\tt
EisenYeast} data from the {\tt biclust} \proglang{R} package \citep{biclust}.
The data consist of 80 experiments involving the
Saccharomyces Cerevisiae yeast measuring gene expression data across 6,221
Affymetrix probes, represented as an $80\times 6221$ real-valued matrix.  The
example finds all pairs of columns (gene expression measurements) that meet or exceed a
Pearson correlation value of 0.95.  We use a projected dimension $p=10$
in {\tt tcor()}.
\begin{small}
<<eval=TRUE,echo=TRUE>>=
library(tcor)
library(doMC)     # local parallel computation
library(biclust)  # for the BicatYeast data
data(EisenYeast)  # in gene by sample orientation
A = t(EisenYeast) # to correlate genes to genes
threshold = 0.95

@@

<<echo=FALSE, results=tex>>=
require(xtable)
ans = data.frame(Method=c("tcor","brute force"), `Wall clock time (s)`=c(t1, t2), `Peak excess memory use (MB)`=c(m1, m2), check.names=FALSE)
print(
    xtable(ans,
        caption = "Gene expression results, threshold=0.95", label = "EisenYeast", caption.placement = "top")
  ,size="small", table.placement="ht", include.rownames=FALSE)
@

@@

Performance benefits from the {\tt tcor()} function become apparent with more
data.  The following example uses calculated beta values from the Illumina
Human Methylation450 BeadChip of 80 Adenoid Cystic Carcinoma (ACC) tumor
samples obtained from the Broad GDAC Cancer Genome Atlas dashboard \citep{gdac}.
See the \proglang{R} package implementation \citep{sup} for examples of
efficiently converting the downloaded sample methylation data into an R matrix.
We chose these data for their public availability as a general example of
computing thresholded correlation among large numbers of vectors,
rather than for any specific biological meaning.

We removed 91,563 columns with all constant or missing values from the data,
leaving an $80\times 394014$ matrix. The example computes pairs of methylation
beta value column vectors with correlation values of 0.99 or more. We used a
{\tt tcor} projection dimension of 10 and the same performance measurement
methodology used in the first example. Despite its low rank and modest data
size, the problem is too large to compute using a na\"ive brute force
correlation method.  (We remark that the matrix multiplication central to the
brute force implementation can be decomposed into a sequence of smaller steps
to fit within given memory constraints. However, the brute force algorithm
would still need to evaluate all of the nearly 160 billion possible vector pair
correlation values.)

@@

We present a simple algorithm and reference \proglang{R} implementation based on a
truncated singular value decomposition (SVD) that efficiently computes
thresholded subsets of large correlation and distance matrices.  We show that
the natural connection between data covariance and the SVD can result in
substantial computational and memory savings, and illustrate these savings with
a few genomics data experiments.  Moreover, the main computational work in the
algorithm is easily parallelized and deployed across computational
clusters.
